Modelling the Retinotectal projection - can competition provide a stopping mechanism?

### 2021-07-05 paper/paper.tex
WIP on paper

One reason for the success of the study of the retinotectal projection is its
capacity to be experimentally manipulated. In some non-mammalian species, both
the retina and the tectum can be partially ablated, or even physically
reorganised in-vivo, after which axons regrow to restore the order and
function of the system for the individual animal. This manipulability was
exploited in influential work by R. W. Sperry and co-workers during the
mid-twentieth century, leading to Sperry's 1963 summary of
the \emph{chemoaffinity theory}~\cite{sperry_chemoaffinity_1963} which
proposes the existence of morphogenetic gradients that guide axons to their
destination. The chemoaffinity theory was given robust support by the
discovery of the ephrin ligands and their receptors~\cite{first_ephrin_paper}
which have been shown to form into graded expression fields in the
retina~\cite{ephrin_retina}, tectum~\cite{ephrin_tectum} as well as in other
sensory systems, such as the somatosensory
system~\cite{vanderhaegen_as_in_elife_paper}. The Ephrin ligands have a clear
effect on axonal outgrowth, as shown in in-vitro~\cite{one} and
in-vivo~\cite{another} studies.
%
% A-P:
%
% EphA and ephrinAs: Anteroposterior axis of retiontectal projection
% (Nasal-Temporal on retina; Anteroposterior on tectum).
%
% RGM (tectum) Neogenin (retina)
%
% D-V:
%
% EphB and ephrin B: Dorsoventral on retina; dorsoventral on tectum.
%
% En-2: Repels temporal growth cones of xenopus ret-tec axons and attracts
% nasal ones. Brunet et all, Nature, 2005
%
% Ryk/Fz (retina) and Wnt3 (tectum) (Flanahan, 2006)
%

@@

Our model consists of agents, representing axonal growth cones, originating
from a square retina and moving upon a square tectum. Each agent carries a set
of 4 receptors, $r$, and 4 ligands, $l$, which it expresses at differing
levels depending upon the soma's origin on the retina. The expression levels
of the receptors and ligands form gradients across the retinal
surface. Receptors are arranged in orthogonal pairs with the gradient of $r_0$
being orthogonal to that of $r_1$. $r_2$, whose gradient is opposite to $r_0$
is orthogonal to $r_3$. Ligands are also arranged in orthogonal pairs, with
$l_0$ opposing $r_0$ (as $r_0$ increases, $l_0$ decreases) and $l_1$
orthogonal to $l_0$ and opposing $r_1$. While we do not explicitly name $r_0$,
$r_1$, etc.~as \emph{EphA}, \emph{EphB}, the suggestion is that $r$ includes
EphA, EphB and Ryk~\cite{schmitt_wntryk_2006} receptors and that $l$ includes
the ephrin-A, ephrin-B and Wnt3~\cite{schmitt_wntryk_2006} ligands.
%
The tectum expresses the same sets of receptors and ligands, also in
orthogonal pairs of gradients, though not necessarily with the same expression
function. Figure~\ref{f:1} shows colour maps of receptor and ligand
expression, in this case experssion maps are linear.

@@

%
% BIBLIOGRAPHY
%
\selectlanguage{English}
\bibliographystyle{abbrvnotitle}
\bibliography{RetinoTectal}

>>>>

One reason for the success of the study of the retinotectal projection is its
capacity to be experimentally manipulated. In some non-mammalian species, both
the retina and the tectum can be partially ablated, or even physically
reorganised in-vivo, after which axons regrow to restore the order and
function of the system for the individual animal. This manipulability was
exploited in influential work by R. W. Sperry and co-workers during the
mid-twentieth century, leading to Sperry's 1963 summary of
the \emph{chemoaffinity theory}~\cite{sperry_chemoaffinity_1963} which
proposes the existence of morphogenetic gradients that guide axons to their
destination. The chemoaffinity theory was given robust support by the
discovery of the ephrin ligands and their receptors~\cite{cheng_complementary_1995,drescher_vitro_1995}
which have been shown to form into graded expression fields in the
retina~\cite{braisted_graded_1997}, tectum~\cite{braisted_graded_1997,feldheim_genetic_2000} as well as in other
sensory systems, such as the somatosensory
system~\cite{vanderhaeghen_mapping_2000}. The Ephrin ligands have a clear
effect on axonal outgrowth, as shown in in-vitro~\cite{cheng_complementary_1995,drescher_vitro_1995,hansen_retinal_2004} and
in-vivo~\cite{rodger_transient_2000,mann_topographic_2002,hindges_ephb_2002} studies.
%
% A-P:
%
% EphA and ephrinAs: Anteroposterior axis of retiontectal projection
% (Nasal-Temporal on retina; Anteroposterior on tectum).
%
% RGM (tectum) Neogenin (retina)
%
% D-V:
%
% EphB and ephrin B: Dorsoventral on retina; dorsoventral on tectum.
%
% En-2: Repels temporal growth cones of xenopus ret-tec axons and attracts
% nasal ones. Brunet et all, Nature, 2005
%
% Ryk/Fz (retina) and Wnt3 (tectum) (Flanahan, 2006)
%

@@

\subsection{Receptor and ligand expression}

Our model was inspired by the agent-based model of \cite{simpson_simple_2011}
and, in common with that work, consists of agents representing axonal growth
cones, originating from a square retina and moving upon a square tectum. In
our model, each retinal ganglion cell projects $n=8$ growth cone agents which
carry a set of 4 receptors, $r$, and 4 ligands, $l$, expressed at levels
determined by the cell's location on the retina. The expression levels of the
receptors and ligands vary with respect to the cell's position on the retinal
surface. Each receptor or ligand varies with respect only to one dimension.
Receptor expression gradients are arranged in orthogonal pairs with the
gradient of $r_0$ being orthogonal to that of $r_1$. $r_2$, whose gradient is
opposite to $r_0$, is orthogonal to $r_3$. Ligands are also arranged in
orthogonal pairs, with $l_0$ opposing $r_0$ (as $r_0$ increases, $l_0$
decreases) and $l_1$ orthogonal to $l_0$ and opposing $r_1$.

Because there is evidence that EphA and EphB receptors are expressed in
exponentially increasing
patterns \cite{reber_relative_2004,feldheim_genetic_2000,brown_topographic_2000,koulakov_stochastic_2004},
we use an exponential form for retinal receptor expressions, in common with
other modelling
studies~\cite{reber_relative_2004,koulakov_stochastic_2004,simpson_simple_2011}.
%
The tectum expresses the same sets of receptors and ligands, also in
orthogonal pairs of gradients. In this work, we did not consider the effect of
reverse signalling from tectal receptors to RGC ligands so we only modelled
tectal ligand expression. Although
several studies model tectal ligand expression with exponential functions~\cite{koulakov_stochastic_2004}, the
experimental evidence for ligand expression is more ambiguous (Supp. Fig
X). Correspondingly, we leave open the possibility that ligand expression may
be best modelled by exponential, linear or logarithmic functions.

While we do not
explicitly name $r_0$, $r_1$, etc.~as \emph{EphA}, \emph{EphB}, the suggestion
is that $r$ includes EphA, EphB and Ryk~\cite{schmitt_wntryk_2006} receptors
and that $l$ includes the ephrin-A, ephrin-B and
Wnt3~\cite{schmitt_wntryk_2006} ligands.

Figure~\ref{f:1} shows colour maps of receptor and ligand
expression, in this case tectal ligand expression maps are linear.

\begin{figure}
\includegraphics[width=\linewidth]{./images/expressions_fig.png}
\caption{Retinal and tectal gene expression.
%
\textbf{A} Retinal receptor expression. Four receptors are expressed and
displayed here in two dual-colour maps. $r_0$ and $r_1$ are red and
green;$r_2$ and $r_3$ are cyan and magenta. $r_0$ increases as an
exponential function of $x$; $r_1$ increases with an exponential function of
$y$. $r_2$ and $r_3$ have the opposite sense. The expression is modelled with
exponential functions because there is ample evidence that shows that retinal
receptor expressions follow this form.
%
\textbf{B} Retinal ligand expression. May not be included.
%
\textbf{C} Tectal receptor expression. May not be included.
%
\textbf{D} Tectal ligand expression. $l_0$ and $l_1$ are red and green; $l_2$
and $l_3$ are cyan and magenta. These expressions are shown in a linear form,
as there is less evidence to determine whether tectal ligand expression is
exponential, linear or logarithmic in form.
}
\label{f:1}
\end{figure}

\subsection{Signalling}

This is where I really want to say ``the only interactions in the model are
via forward (ligand-activates-receptor) and reverse
(receptor-activates-ligand) signalling''.

% Hill Equation! https://en.wikipedia.org/wiki/Hill_equation_(biochemistry)

@@

%
% BIBLIOGRAPHY
%
\selectlanguage{English}
\bibliographystyle{apalike}
\bibliography{RetinoTectal}
